---
title: "Peer pressure: Are the fitness benefits of animal personality traits affected by conspecific density?"
subtitle: "Ho et al. (2023)"
date: "Jul 7, 2023"
output: 
  html_document: 
    fig_caption: yes
    theme: flatly
---

```{r}
library(tidyverse)
library(RMySQL)
library(lme4)
```

These analyses are largely based on the code that Matthew Whalen wrote in May 2022. I tweaked the code as needed as we finalized the models and analyses.

# Data Cleaning

`scripts/r/data-cleaning.R`

We read in raw personality data from the master file (`data/personality-master-updated.csv`) and filtered for juveniles. We also ran filters to exclude:

-   Records that indicated that they should be excluded

-   Personality data scored by SWK due to GC experiment

-   Any re-trials (i.e., only included the first trial for each individual)

-   Squirrels from any grid other than **JO, KL, and SU**

The model in `scripts/r/model-training.R` was used to predict and impute as many missing parturition dates as possible.

```{r}
personality <- read.csv("data/personality-mrw-imputed.csv")

glimpse(personality)
```

After this, we get survival data to autumn and overwinter (200 days) in `scripts/r/survival_data.R` by pulling census data from the database.

I also added in a column to indicate mast years (2005 and 2019) and non-mast years.
