---
title: "Peer pressure: Are the fitness benefits of animal personality traits affected by conspecific density?"
subtitle: "Ho et al. (2023)"
date: "Jul 7, 2023"
output: 
  html_document: 
    fig_caption: yes
    theme: lumen
    highlight: tango
---

```{r, include = FALSE}
options(tidyverse.quiet = TRUE)
library(tidyverse)
library(RMySQL)
library(ade4) #for PCA
library(ggplot2)
library(lme4) #for glmers
library(car)
library(DHARMa)
```

These analyses are largely based on the code that Matthew Whalen wrote in May 2022. I tweaked the code as needed as we finalized the models and analyses.

# Data cleaning

`scripts/r/data-cleaning.R`

We read in raw personality data from the master file (`data/personality-master-updated.csv`) and filtered for juveniles. We also ran filters to exclude:

-   Records that indicated that they should be excluded

-   Personality data scored by SWK due to GC experiment

-   Any re-trials (i.e., only included the first trial for each individual)

-   Squirrels from any grid other than **JO, KL, and SU**

The model in `scripts/r/model-training.R` was used to predict and impute as many missing parturition dates as possible.

```{r}
personality <- read.csv("data/personality-mrw-imputed.csv")

glimpse(personality)
```

After this, we get survival data to autumn and overwinter (200 days) in `scripts/r/survival_data.R` by pulling census data from the database.

I also added in a column to indicate mast years (2005 and 2019) and non-mast years and to indicate treatment ('rattle' for JO, 'control' for KL and SU).

# Preparing data for analysis

First, we standardize growth rate and part dates by grid-year combination.

```{r}
personality = read_csv('data/personality-mrw-survival.csv', show_col_types = FALSE) %>% 
  mutate(survival = as.integer(survived_200d)) %>% 
  group_by(grid, year) %>% 
  mutate(growth_sc = scale(growth, scale = T, center = T)[,1],
         part_sc = scale(part, scale = T, center = T)[,1]) %>% 
  ungroup() 
```

Next, we conduct PCA on the OFT/MIS data to get loadings for activity and aggression.

```{r}

personality[is.na(personality$oft_duration),
            "oft_duration"] <- 450.000
personality[is.na(personality$mis_duration),
            "mis_duration"] <- 300.000

aggression = personality %>% 
  select(front, back, attack, attacklatency, approachlatency, mis_duration) %>% 
  mutate(across(everything(), ~if_else(is.na(.x), median(.x, na.rm = TRUE), .x))) %>%
  mutate(front = front/mis_duration,
         back = back/mis_duration,
         attack = attack/mis_duration,
         attacklatency = attacklatency/mis_duration,
         approachlatency = approachlatency/mis_duration) %>%
  select(-mis_duration)

pca_agg = prcomp(aggression, scale = TRUE, center = TRUE)
mis1 = predict(pca_agg)[,1]

activity = personality %>% 
  select(walk, still, hang, jump, chew, hole, groom, oft_duration) %>% 
  mutate(across(everything(), ~if_else(is.na(.x), median(.x, na.rm = TRUE), .x)))%>%
  mutate(walk = walk/oft_duration,
         still = still/oft_duration,
         hang = hang/oft_duration,
         jump = jump/oft_duration,
         chew = chew/oft_duration,
         hole = hole/oft_duration,
         groom = groom/oft_duration) %>%
  select(-oft_duration)

pca_act = prcomp(activity, scale = TRUE, center = TRUE)
oft1 = predict(pca_act)[,1]
```

The loadings came out reversed, so we have to multiply them by -1.

```{r}
personality$oft1 = unlist(oft1 * -1)
personality$mis1 = unlist(mis1 * -1)
```

Here are the plotted PCA loadings with a 95% ellipse.

```{r}
ggplot(personality, aes(oft1, mis1)) +
  geom_point(size = 0.75) +
  stat_ellipse(level = 0.95) +
  labs(x = "Activity (PC1)",
       y = "Aggression (PC1)") +
  theme_minimal() +
  theme(plot.margin = margin(20,20,20,20))
```

We scale grid density and age at time of trial globally, and activity (*oft1*) and aggression (*mis1*) by grid-year combination.

```{r}
personality <- personality %>%
  mutate(grid_density = scale(grid_density, scale = T, center = T)[,1],
         age_sc = scale(age_at_trial, scale = T, center = T)[,1]) %>%
  group_by(grid, year) %>%
  mutate(oft1 = scale(oft1, scale = T, center = T)[,1],
         mis1 = scale(mis1, scale = T, center = T)[,1]) %>%
  ungroup() %>%
  mutate(treatment = factor(treatment))
```

# Generalized mixed-effects models

## Survival to autumn

Here's the model for survival to autumn. Fixed effects include:

-   Activity

-   Aggression

-   Part date

-   Growth rate

-   Mast/non-mast year

-   Treatment

-   Grid density

    -   Interactions with activity, aggression, part date, and growth rate

Random effects include:

-   Year (of birth)

-   Litter ID

```{r}
dat = personality %>% 
  mutate(across(c(year, dam_id, litter_id, grid, mastyear), as_factor)) 

survival_to_autumn = glmer(made_it ~
                             oft1*mis1*grid_density +
                             part_sc*grid_density +
                             growth_sc*grid_density +
                             mastyear +
                             treatment +
                             (1|year) +
                             (1|litter_id), 
                           data = dat,
                           na.action = 'na.omit',
                           family = binomial(link = "logit"),
                           control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e8)))
```

### GLM

```{r}
car::Anova(survival_to_autumn)
```

-   Significant effect of mast year

-   Significant interaction between growth rate and grid density

-   Near-significant interaction between activity and aggression

-   Near-significant interactions between activity, aggression, and grid density

### Model effects

```{r}
summary(survival_to_autumn)
```

Here, we see that:

-   Juvs have better survival during mast years

-   More active and aggressive juvs survive better

-   Slower growing juvs survive better under high density conditions

-   There is a near-significant interaction effect between aggression and conspecific density (more aggressive juvs do better under high density conditions)

### Model diagnostics

Checking for any collinearity between predictors, where VIF \> 3 indicates collinearity:

```{r}
vif(survival_to_autumn)
```

All good here.

```{r, results = 'hide'}
simulationOutput = simulateResiduals(survival_to_autumn, plot = F)
residuals(simulationOutput, quantileFunction = qnorm)
```

```{r}
plot(simulationOutput)
```

No deviations here.

## Overwinter survival (200 days)

Here's the model for overwinter survival. All the fixed effects and random effects are the same as that of the survival to autumn model, except with *survived_to_200d* as the response variable instead.

```{r}
dat2 = personality %>% 
  mutate(across(c(year, dam_id, litter_id, grid, mastyear), as_factor))

survival_to_200d = glmer(survived_200d ~ 
                           oft1*mis1*grid_density +
                           part_sc*grid_density +
                           growth_sc*grid_density +
                           mastyear +
                           treatment +
                           (1|year) +
                           (1|litter_id),
                         data = dat2,
                         na.action = 'na.omit',
                         family = 'binomial',
                         control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e8)))
```

### GLM

```{r}
car::Anova(survival_to_200d)
```

-   Significant effect of mast year

-   Significant interaction between activity and density

-   Near-significant interaction between activity and aggression.

### Model effects

```{r}
summary(survival_to_200d)
```

Here, we see that:

-   Juvs have better survival during mast years

-   More active and aggressive juvs survive better

-   More active juvs survive better under high density conditions

### Model diagnostics

Checking for any collinearity between predictors, where VIF \> 3 indicates collinearity:

```{r}
vif(survival_to_200d)
```

All good here.

```{r, results='hide'}
simulationOutput = simulateResiduals(survival_to_200d, plot = F)
residuals(simulationOutput, quantileFunction = qnorm)
```

```{r}
plot(simulationOutput)
```

No deviations here either.
